Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-07 _tex/paper2.tex
Built site for gh-pages

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates
\citep[pp.~2]{Perron_et_al_2015}. First, it critiques the heavy reliance
on Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
extension, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the disconnect
between CJ's trait measurement and hypothesis testing processes.
Although separating these processes simplifies the analysis of CJ data,
it may also undermine the reliability of various statistical results
derived from these processes.

@@

In its most general form, Thurstone's theory addresses pairwise
comparisons wherein a single judge evaluates multiple stimuli
\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the discriminal process for any
given stimulus forms a Normal distribution along the trait continuum
\citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
known as the \emph{modal discriminal process}, indicates the stimulus
position on this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep[pp.~251]{Bramley_2008}. This suggests that pairwise comparison
outcomes depend on the relative distance between stimuli, not their
absolute positions on the continuum. Indeed, the theory assumes that the
difference between the underlying discriminal processes of the stimuli,
referred to as the \emph{discriminal difference}, determines the
observed dichotomous outcome. Furthermore, the theory assumes that
because the individual discriminal processes form a Normal distribution
on the continuum, the discriminal difference will also conform to a
Normal distribution \citep[pp.~452]{Andrich_1978}. In this distribution,
the mode (mean) represents the relative separation between the stimuli,
and its dispersion indicates the variability of that separation.

@@

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement, and without providing clear guidance on
how to use its measurement scores for hypothesis testing. Specifically,
he cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, as the case imposes the most
extensive set of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
(see Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's
primary goal was to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
this continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not address
how to use the resulting measurement scores for hypothesis testing.
Thus, given these limitations, it is surprising that CJ research has
predominantly relied on Case V to measure different traits, raising
significant concerns about the reliability and validity of such
measurements in contexts where the case's assumptions may not hold
\citep[pp.~677]{Kelly_et_al_2022}. Furthermore, although the CJ practice
has attempted to address the issue of hypothesis testing by using the
scores' point estimates or their transformations, a critical question
remains: Is this approach suitable for addressing these inquiries?

@@

As previously discussed, Case V remains the most widely used model in CJ
literature. This preference primarily stems from the BTL model, which
provides a simplified statistical representation of the case. The BTL
model mirrors the assumptions of Case V, with one notable distinction:
while Case V assumes a Normal distribution for the stimuli' discriminal
processes, the BTL model uses the more mathematically tractable Logistic
distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on the model's estimation or interpretation because the
discriminal process scale is arbitrary up to a non-monotonic
transformation
\citetext{\citealp[pp.~16]{vanderLinden_et_al_2017_I}; \citealp{McElreath_2021}}.
Furthermore, this limited impact is supported by the fact that the
Normal and Logistic distributions exhibit analogous statistical
properties, differing only by a scaling factor of approximately \(1.7\).

However, Thurstone acknowledged that some assumptions of Case V could be
problematic when researchers assess complex traits or heterogeneous
stimuli \citep[pp.~376]{Thurstone_1927a}. Thus, given that modern CJ
applications often involve such traits and stimuli, two key assumptions
of Case V, and by extension, the BTL model, may not always hold in
theory or practice. These assumptions are the equal dispersion and zero
correlation between stimuli.

@@

In experimental practice, however, the thought experiment occurs in
reverse. Researchers first observe the comparison outcome and then use
the BTL model to infer the discriminal difference between stimuli and
their respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
Consequently, the outcome's ability to reflect \emph{true} differences
between stimuli largely depends on the validity of the model's
assumptions \citep[pp.~150]{Kohler_et_al_2019}, in this case, the
assumption of equal dispersions. For instance, when the assumption
accurately captures the complexity of the data, the BTL model estimates
a discriminal difference distribution that accurately represents the
\emph{true} discriminal difference between the texts. This scenario is
illustrated in Figure~\ref{fig-dispersion}, when the model's discriminal
difference distribution aligns with the \emph{true} discriminal
difference distribution, represented by the thick continuous line
corresponding to \(\sigma_{B}-\sigma_{A}=0\). The accuracy of this
discriminal difference then ensures reliable estimates for the texts'
discriminal processes.

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to capture the complexity of modern CJ data. In
particular, the assumption may not hold when researchers assess complex
traits or heterogeneous stimuli
\citetext{\citealp[pp.~376]{Thurstone_1927a}; \citealp[pp.~273]{Bramley_2008}; \citealp[pp.~678]{Kelly_et_al_2022}},
as these traits and stimuli can introduce judgment discrepancies due to
their unique characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide indications of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.
These misfit texts may exhibit larger discriminal dispersions or
represent outliers --texts with distinctive characteristics that deviate
markedly from the rest of the sample \citep{Grubbs_1969}. In either
case, comparing misfit texts results in more judgment discrepancies, yet
the BTL model neither accounts for these cases nor offers any means to
address them.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent outlying observations,
the common CJ practice of excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
may unintentionally discard valuable information, introducing bias into
the trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}.
The direction and magnitude of these biases remain unpredictable, as
they depend on which stimuli researchers exclude from the analysis.

@@

Thurstone assumed that stimuli were uncorrelated because judges' biases,
arising from two opposing and equally weighted effects occurring during
the pairwise comparisons, canceled each other out
\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} provided a
mathematical demonstration of this cancellation using the BTL model
under the assumption of discriminal processes with additive biases.
However, evidence from the CJ literature indicates that the assumption
of zero correlation does not hold in practice in at least two scenarios:
when intricate aspects of multidimensional, complex traits or
heterogeneous stimuli influence judges' perceptions or when additional
hierarchical structures are relevant to the stimuli.

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on various intricate characteristics of the
stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, when assessed, are unlikely to be
equally weighted or opposing. As a result, they may exert an uneven
influence on judges' perceptions, creating biases that resist
cancellation. For example, this could occur when a judge assessing the
argumentative quality of a text places more weight on its grammatical
accuracy than other judges, thereby favoring texts with fewer errors but
weaker arguments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases could introduce dependencies between the
stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}. While direct
evidence for this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} and \citet{vanDaal_et_al_2016} demonstrate
the presence of such biases, supporting the notion that the factors
influencing pairwise comparisons may not always cancel out.

In the second scenario, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli, a statistical phenomenon known as clustering
\citep[pp.~86]{Everitt_et_al_2010}. For instance, when multiple texts
are produced by the same individual, these texts are likely to share
common features, such as writing style or general quality. Although the
CJ literature acknowledges the existence of such hierarchical
structures, the statistical approaches to account for this additional
source of dependence have been insufficient. For instance, when CJ data
incorporates multiple samples of stimuli from the same individuals,
researchers frequently rely on (averaged) point estimates of the BTL
scores to conduct subsequent analyses and tests at the individual level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. Specifically,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability \citep[pp.~341,
482]{Ackerman_1989, Hoyle_et_al_2023}. These inaccuracies can result in
spurious conclusions about the discriminal differences
\citep[pp.~370]{McElreath_2020} and, by extension, the underlying
discriminal processes of the stimuli. This issue is illustrated in
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the BTL scores follows the thick continuous line
\((\rho = 0)\), while the \emph{true} discriminal difference follows any
discontinuous line where \(\rho \neq 0\).

Finally, as discussed in the previous section, removing \emph{misfit}
judges based risks discarding valuable information and even introduce
bias into the trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases remain unpredictable because they depend on
which judges researchers exclude from the analysis. \emph{Misfit judges}
are those whose evaluations deviate substantially from the shared
consensus due to the unique characteristics of either the stimuli or the
judges themselves
\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.

@@

Building on the previous section, it is clear that, researchers
typically rely on the BTL model to measure a trait and place the
compared stimuli along its continuum \citep[pp.~269]{Thurstone_1927b}.
Additionally, the CJ literature shows that researchers frequently use
point estimates of BTL scores or their transformations to conduct
further analyses or hypothesis tests. For example, researchers have used
these scores to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty (measurement error). Ignoring this uncertainty can bias the
analysis and reduce the precision of hypothesis tests. The direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Furthermore, the reduced precision in hypothesis tests diminishes their
statistical power, increasing the likelihood of committing type-I or
type-II errors \citep{McElreath_2020}.
Figure~\ref{fig-measurement_error} illustrates these issues,
demonstrating how neglecting measurement error \((\sigma_{T})\) by
relying only on outcome averages can reduce the precision of a
predictor's estimated effect.

@@

In aggregate, the heavy reliance on Thurstone's Case V assumptions in
the statistical analysis of comparative data can compromise the
reliability of trait estimates. This overreliance may also undermine
their validity \citep[pp.~2]{Perron_et_al_2015}, particularly when
coupled with the disconnect between the trait measurement and hypothesis
testing processes. However, the structural approach to causal inference
can address these issues by offering a systematic and integrated
framework that strengthens measurement reliability and validity while
enhancing the statistical accuracy of hypothesis tests.

@@

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. The approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ experiments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests \citep[pp.~4]{Schuessler_et_al_2023}.
\emph{Identification analysis} helps researchers to determine whether an
estimator can accurately compute an estimand based solely on its
(causal) assumptions, regardless of random variability
\citep[pp.~4]{Schuessler_et_al_2023}. Here, \emph{estimands} represent
the specific quantities researchers aim to determine
\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
functions that transform data into an estimate, while \emph{estimates}
are the numerical values approximating the estimand
\citep{Neal_2020, Everitt_et_al_2010}.

@@

However, Section~\ref{sec-theory-issues} presents compelling evidence
that Thurstone's Case V, and by extension the BTL model, suffers from
several statistical and measurement limitations. These limitations
hinder the model's ability to identify various estimands relevant to CJ
inquiries, including the one described in the example. Identification is
crucial because it is a necessary condition for ensuring consistent
estimators. \emph{Consistency} refers to the property of an estimator
whose estimates converge to the ``true'' value of the estimand as the
data size approaches infinity \citep{Everitt_et_al_2010}. Without
identification, consistency cannot be achieved, even with ``infinite''
and error-free data. Thus, deriving meaningful insights from finite data
becomes impossible \citep[pp.~5]{Schuessler_et_al_2023}.

Fortunately, SCMs and DAGs support identification analysis through two
key advantages. First, regardless of complexity, they can represent
various causal structures using only five fundamental building blocks
\citep[see][]{Neal_2020, McElreath_2020}. This feature allows
researchers to decompose complex structures into manageable components,
facilitating their analysis. Second, they depict causal relationships in
a non-parametric way. This flexibility enables feasible identification
strategies without requiring specification of the types of variables,
the functional forms relating them, or the parameters of those
functional forms \citep[pp.~35]{Pearl_et_al_2016}.

@@

\[
\begin{aligned}
  O^{cp}_{hiabjk} & := f_{O}(D_{hiabjk}) \\
  D_{hiabjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
\end{aligned}
\]

@@

\caption{\label{fig-cj04}Conceptual-population model, vectorized form.}

@@

@@

\caption{\label{fig-cj14}Sample-comparison model including sample and
comparison mechanisms}

@@

>>>>

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates
\citep{Perron_et_al_2015}. First, it critiques the heavy reliance on
Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
extension, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the disconnect
between CJ's trait measurement and hypothesis testing processes.
Although separating these processes simplifies the analysis of CJ data,
it may also undermine the reliability of various statistical results
derived from these processes.

@@

In its most general form, Thurstone's theory addresses pairwise
comparisons wherein a single judge evaluates multiple stimuli
\citep{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the discriminal process for any
given stimulus forms a Normal distribution along the trait continuum
\citep{Thurstone_1927b}. The mode (mean) of this distribution, known as
the \emph{modal discriminal process}, indicates the stimulus position on
this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep{Bramley_2008}. This suggests that pairwise comparison outcomes
depend on the relative distance between stimuli, not their absolute
positions on the continuum. Indeed, the theory assumes that the
difference between the underlying discriminal processes of the stimuli,
referred to as the \emph{discriminal difference}, determines the
observed dichotomous outcome. Furthermore, the theory assumes that
because the individual discriminal processes form a Normal distribution
on the continuum, the discriminal difference will also conform to a
Normal distribution \citep[pp.~452]{Andrich_1978}. In this distribution,
the mode (mean) represents the relative separation between the stimuli,
and its dispersion indicates the variability of that separation.

@@

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement, and without providing clear guidance on
how to use its measurement scores for hypothesis testing. Specifically,
he cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, as the case imposes the most
extensive set of simplifying assumptions
\citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
this continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not address
how to use the resulting measurement scores for hypothesis testing.
Thus, given these limitations, it is surprising that CJ research has
predominantly relied on Case V to measure different traits, raising
significant concerns about the reliability and validity of such
measurements in contexts where the case's assumptions may not hold
\citep{Kelly_et_al_2022}. Furthermore, although the CJ practice has
attempted to address the issue of hypothesis testing by using the
scores' point estimates or their transformations, a critical question
remains: Is this approach suitable for addressing these inquiries?

@@

As previously discussed, Case V remains the most widely used model in CJ
literature. This preference primarily stems from the BTL model, which
provides a simplified statistical representation of the case. The BTL
model mirrors the assumptions of Case V, with one notable distinction:
while Case V assumes a Normal distribution for the stimuli' discriminal
processes, the BTL model uses the more mathematically tractable Logistic
distribution \citep{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on the model's estimation or interpretation because the
discriminal process scale is arbitrary up to a non-monotonic
transformation \citep{vanderLinden_et_al_2017_I, McElreath_2021}.
Furthermore, this limited impact is supported by the fact that the
Normal and Logistic distributions exhibit analogous statistical
properties, differing only by a scaling factor of approximately \(1.7\).

However, Thurstone acknowledged that some assumptions of Case V could be
problematic when researchers assess complex traits or heterogeneous
stimuli \citep{Thurstone_1927a}. Thus, given that modern CJ applications
often involve such traits and stimuli, two key assumptions of Case V,
and by extension, the BTL model, may not always hold in theory or
practice. These assumptions are the equal dispersion and zero
correlation between stimuli.

@@

In experimental practice, however, the thought experiment occurs in
reverse. Researchers first observe the comparison outcome and then use
the BTL model to infer the discriminal difference between stimuli and
their respective discriminal processes \citep{Thurstone_1927a}.
Consequently, the outcome's ability to reflect \emph{true} differences
between stimuli largely depends on the validity of the model's
assumptions \citep{Kohler_et_al_2019}, in this case, the assumption of
equal dispersions. For instance, when the assumption accurately captures
the complexity of the data, the BTL model estimates a discriminal
difference distribution that accurately represents the \emph{true}
discriminal difference between the texts. This scenario is illustrated
in Figure~\ref{fig-dispersion}, when the model's discriminal difference
distribution aligns with the \emph{true} discriminal difference
distribution, represented by the thick continuous line corresponding to
\(\sigma_{B}-\sigma_{A}=0\). The accuracy of this discriminal difference
then ensures reliable estimates for the texts' discriminal processes.

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to capture the complexity of modern CJ data. In
particular, the assumption may not hold when researchers assess complex
traits or heterogeneous stimuli
\citep{Thurstone_1927a, Bramley_2008, Kelly_et_al_2022}, as these traits
and stimuli can introduce judgment discrepancies due to their unique
characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide indications of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citep{Pollitt_2004, Pollitt_2012b}. These misfit texts may exhibit
larger discriminal dispersions or represent outliers --texts with
distinctive characteristics that deviate markedly from the rest of the
sample \citep{Grubbs_1969}. In either case, comparing misfit texts
results in more judgment discrepancies, yet the BTL model neither
accounts for these cases nor offers any means to address them.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent outlying observations,
the common CJ practice of excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
may unintentionally discard valuable information, introducing bias into
the trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}.
The direction and magnitude of these biases remain unpredictable, as
they depend on which stimuli researchers exclude from the analysis.

@@

Thurstone assumed that stimuli were uncorrelated because judges' biases,
arising from two opposing and equally weighted effects occurring during
the pairwise comparisons, canceled each other out
\citep{Thurstone_1927b}. \citet{Andrich_1978} provided a mathematical
demonstration of this cancellation using the BTL model under the
assumption of discriminal processes with additive biases. However,
evidence from the CJ literature indicates that the assumption of zero
correlation does not hold in practice in at least two scenarios: when
intricate aspects of multidimensional, complex traits or heterogeneous
stimuli influence judges' perceptions or when additional hierarchical
structures are relevant to the stimuli.

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on various intricate characteristics of the
stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, when assessed, are unlikely to be
equally weighted or opposing. As a result, they may exert an uneven
influence on judges' perceptions, creating biases that resist
cancellation. For example, this could occur when a judge assessing the
argumentative quality of a text places more weight on its grammatical
accuracy than other judges, thereby favoring texts with fewer errors but
weaker arguments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases could introduce dependencies between the
stimuli \citep{vanderLinden_et_al_2017_II}. While direct evidence for
this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} and \citet{vanDaal_et_al_2016} demonstrate
the presence of such biases, supporting the notion that the factors
influencing pairwise comparisons may not always cancel out.

In the second scenario, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli, a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. For instance, when multiple texts are
produced by the same individual, these texts are likely to share common
features, such as writing style or general quality. Although the CJ
literature acknowledges the existence of such hierarchical structures,
the statistical approaches to account for this additional source of
dependence have been insufficient. For instance, when CJ data
incorporates multiple samples of stimuli from the same individuals,
researchers frequently rely on (averaged) point estimates of the BTL
scores to conduct subsequent analyses and tests at the individual level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. Specifically,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability
\citep{Ackerman_1989, Hoyle_et_al_2023}. These inaccuracies can result
in spurious conclusions about the discriminal differences
\citep{McElreath_2020} and, by extension, the underlying discriminal
processes of the stimuli. This issue is illustrated in
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the BTL scores follows the thick continuous line
\((\rho = 0)\), while the \emph{true} discriminal difference follows any
discontinuous line where \(\rho \neq 0\).

Finally, as discussed in the previous section, removing \emph{misfit}
judges based risks discarding valuable information and even introduce
bias into the trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases remain unpredictable because they depend on
which judges researchers exclude from the analysis. \emph{Misfit judges}
are those whose evaluations deviate substantially from the shared
consensus due to the unique characteristics of either the stimuli or the
judges themselves
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.

@@

Building on the previous section, it is clear that, researchers
typically rely on the BTL model to measure a trait and place the
compared stimuli along its continuum \citep{Thurstone_1927b}.
Additionally, the CJ literature shows that researchers frequently use
point estimates of BTL scores or their transformations to conduct
further analyses or hypothesis tests. For example, researchers have used
these scores to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty (measurement error). Ignoring this uncertainty can bias the
analysis and reduce the precision of hypothesis tests. The direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citep{Kline_et_al_2023, Hoyle_et_al_2023}. Furthermore, the reduced
precision in hypothesis tests diminishes their statistical power,
increasing the likelihood of committing type-I or type-II errors
\citep{McElreath_2020}. Figure~\ref{fig-measurement_error} illustrates
these issues, demonstrating how neglecting measurement error
\((\sigma_{T})\) by relying only on outcome averages can reduce the
precision of a predictor's estimated effect.

@@

In aggregate, the heavy reliance on Thurstone's Case V assumptions in
the statistical analysis of comparative data can compromise the
reliability of trait estimates. This overreliance may also undermine
their validity \citep{Perron_et_al_2015}, particularly when coupled with
the disconnect between the trait measurement and hypothesis testing
processes. However, the structural approach to causal inference can
address these issues by offering a systematic and integrated framework
that strengthens measurement reliability and validity while enhancing
the statistical accuracy of hypothesis tests.

@@

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. The approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ experiments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests. \emph{Identification analysis} helps researchers to
determine whether an estimator can accurately compute an estimand based
solely on its (causal) assumptions, regardless of random variability
\citep[pp.~4]{Schuessler_et_al_2023}. Here, \emph{estimands} represent
the specific quantities researchers aim to determine
\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
functions that transform data into an estimate, while \emph{estimates}
are the numerical values approximating the estimand
\citep{Neal_2020, Everitt_et_al_2010}.

@@

However, Section~\ref{sec-theory-issues} presents compelling evidence
that Thurstone's Case V, and by extension the BTL model, suffers from
several statistical and measurement limitations. These limitations
hinder the model's ability to identify various estimands relevant to CJ
inquiries, including the one described in the example. Identification is
crucial because it is a necessary condition for ensuring consistent
estimators. \emph{Consistency} refers to the property of an estimator
whose estimates converge to the ``true'' value of the estimand as the
data size approaches infinity \citep{Everitt_et_al_2010}. Without
identification, consistency cannot be achieved, even with ``infinite''
and error-free data. Thus, deriving meaningful insights from finite data
becomes impossible \citep{Schuessler_et_al_2023}.

Fortunately, SCMs and DAGs support identification analysis through two
key advantages\footnote{These topics are beyond the scope of this study,
  thus, readers seeking a more profound understanding can refer to
  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, and introductory
  books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
  \citet{McElreath_2020} are useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
  \citet{Hernan_et_al_2020}, are recommended.}. First, regardless of
complexity, they can represent various causal structures using only five
fundamental building blocks \citep{Neal_2020, McElreath_2020}. This
feature allows researchers to decompose complex structures into
manageable components, facilitating their analysis. Second, they depict
causal relationships in a non-parametric way. This flexibility enables
feasible identification strategies without requiring specification of
the types of variables, the functional forms relating them, or the
parameters of those functional forms \citep{Pearl_et_al_2016}.

@@

\[
\begin{aligned}
  O^{cp}_{haibjk} & := f_{O}(D_{haibjk}) \\
  D_{haibjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
\end{aligned}
\]

@@

\caption{\label{fig-cj04}Conceptual-population model, initial vectorized
form.}

@@

Overall, the conceptual-population model extends Thurstone's general
form by introducing key innovations to address the limitations discussed
in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
These enhancements include accounting for judges' biases and integrating
hierarchical structural components. Nevertheless, despite its promise of
enhancing measurement accuracy and precision, the model still depends on
the unrealistic assumption that researchers have access to a
\emph{conceptual population}. Since researchers rarely meet this
assumption in practice, they must consider a more realistic scenario.

@@

\caption{\label{fig-cj14}Sample-comparison model, final vectorized form}

@@

Using SCM~\ref{fig-cj15_scm} we finally define the
